---
title: "variables"
author: "YunranChen"
date: "2/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=TRUE, 
               comment=NA, verbose=TRUE, fig.width=6, fig.height=4)
library(corrplot)
library(naniar)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(lme4)
library(jpeg)
library(ggpubr)
library(grid)
library(ggmosaic)
library(stringr)
library(tidyverse)
library(readr)
library(knitr)
library(stringr)
library("lavaan")
theme_set(theme_bw(base_size=12))

```

2001:

About academic: f5 (GPA); f1-f4
Smoking behavior: e3 (most important)+(e4-e10); e11-e14(potential/passive smoking behavior); use e3 to impute e4-e10
drinking behavior: c7, c11, c12, and problem_drinking(maybe also include c4)
Greek life: a5,a6,a8d

background corr>0.1: a2,a7b,a7c(all related to gpa, a7b,a7c also related to smooking)g6,g7,g8,g3,g15,g16(insteresting that mother's drinking behavior influence a lot compared to father's)

background corr>0.05:a1,a2,a3,a4a,a6a,a7b,a7c,a7d,g1,g2,g6,g7,g12a,g17,g18,g3,"g4"  "g5a" "g5b" "g6"  "g14" "g15" "g16"

6.07% missing: 10242/10904
8.63% missing: 9984/10904 if include greek life

```{r load data}
load("dat.rdata")
nrow(dat1997)
dat2001=dat2001 %>% dplyr::select(-coll_id,-serial)
## Study id of the complete cases
ids01=dat2001%>% dplyr::select(study_id,f5,e3,e11:e14,c1,c2,c7,c10)%>%na.omit()%>%pull(study_id)

#with greek life
ids01_wgl=dat2001%>% dplyr::select(study_id,a5,a6,a8d,f5,e3,e11:e14,c1,c2,c7,c10)%>%na.omit()%>%pull(study_id)

# select complete cases
dat_01=dat2001%>%filter(study_id%in%ids01_wgl,f5!=99,f5!=10)
#impute the g3
g3im=dat_01%>%select(study_id,g3a:g3e)%>%gather(.,g3,value,-study_id)%>%na.omit()%>%select(-value)%>%mutate(g3=as.factor(g3)%>%as.numeric())
dat_01=left_join(dat_01,g3im,"study_id")%>%dplyr::select(-g3a,-g3b,-g3c,-g3d,-g3e,-g12b)

## impute the data
# impute e4-e10 using e3
dat_01$e4[dat_01$e3==1]=1
dat_01$e4[dat_01$e3%in%c(2:5)]=2
dat_01$e5[dat_01$e3==1]=0 # 0-represent non-smoker
dat_01$e6[dat_01$e3==1]=0
dat_01$e7[dat_01$e3==1]=1
dat_01$e7[dat_01$e3==2]=1
dat_01$e8[dat_01$e3==1]=1
dat_01$e8[dat_01$e3==2]=1
dat_01$e9[dat_01$e3==1]=1 
dat_01$e10[dat_01$e3==1]=0 # 0-represent non-smoker
# impute c4
dat_01$c4[(dat_01$c1==1)&(dat_01$c2==1)]=0 # 0-represent non frequent drinker
dat_01$c4[(dat_01$c1==1)&(dat_01$c2>=1)&is.na(dat_01$c4)]=1
```

Variables for Drinking 2001 

```{r}
#Use C7, C11, C12, and problem_drinking for DRINKING latent variable
#Use A5, A6, A8d for GREEK latent variable

#Imputing NA's as zero in  C11, C12
dat_01$c11[is.na(dat_01$c11) == TRUE] <- 1
dat_01$c12[is.na(dat_01$c12) == TRUE] <- 1

#Imputing NA's as "None" in c17
dat_01$c17a[is.na(dat_01$c17a) == TRUE] <- 1
dat_01$c17b[is.na(dat_01$c17b) == TRUE] <- 1
dat_01$c17c[is.na(dat_01$c17c) == TRUE] <- 1
dat_01$c17d[is.na(dat_01$c17d) == TRUE] <- 1
dat_01$c17e[is.na(dat_01$c17e) == TRUE] <- 1
dat_01$c17f[is.na(dat_01$c17f) == TRUE] <- 1
dat_01$c17g[is.na(dat_01$c17g) == TRUE] <- 1
dat_01$c17h[is.na(dat_01$c17h) == TRUE] <- 1
dat_01$c17i[is.na(dat_01$c17i) == TRUE] <- 1
dat_01$c17j[is.na(dat_01$c17j) == TRUE] <- 1
dat_01$c17k[is.na(dat_01$c17k) == TRUE] <- 1
dat_01$c17l[is.na(dat_01$c17l) == TRUE] <- 1


#Creating indicators for "3 or more times"
dat_01 <- dat_01 %>% 
  mutate(indic_c17a = ifelse(c17a >=3, 1, 0)) %>% 
  mutate(indic_c17b = ifelse(c17b >=3, 1, 0)) %>%
  mutate(indic_c17c = ifelse(c17c >=3, 1, 0)) %>%
  mutate(indic_c17d = ifelse(c17d >=3, 1, 0)) %>%
  mutate(indic_c17e = ifelse(c17e >=3, 1, 0)) %>%
  mutate(indic_c17f = ifelse(c17f >=3, 1, 0)) %>%
  mutate(indic_c17g = ifelse(c17g >=3, 1, 0)) %>%
  mutate(indic_c17h = ifelse(c17h >=3, 1, 0)) %>%
  mutate(indic_c17i = ifelse(c17i >=3, 1, 0)) %>%
  mutate(indic_c17j = ifelse(c17j >=3, 1, 0)) %>%
  mutate(indic_c17k = ifelse(c17k >=3, 1, 0)) %>%
  mutate(indic_c17l = ifelse(c17l >=3, 1, 0)) %>% 
  mutate(problem_drinking = indic_c17a + indic_c17b + indic_c17c + indic_c17d + indic_c17e + indic_c17f + indic_c17g + indic_c17h + indic_c17i +indic_c17j + indic_c17k + indic_c17l)
```

# final dataset for 2001

About academic: f5 (GPA); f1-f4
Smoking behavior: e3 (most important),e4,e8; e11-e14(potential/passive smoking behavior); use e3 to impute e4-e10
drinking behavior: c7, c11, c12, and problem_drinking(maybe also include c4)
Greek life: a5,a6,a8d

background corr>0.1: a2,a7(all related to gpa, a7b,a7c also related to smooking)g6,g7,g8,g3,g15,g16(insteresting that mother's drinking behavior influence a lot compared to father's)

```{r final2001}
a7im=dat_01%>%select(study_id,a7a:a7d)%>%gather(.,a7,value,-study_id)%>%na.omit()%>%filter(value>0)%>%select(-value)%>%mutate(a7=as.factor(a7))
a7imid=dat_01%>%select(study_id,a7a:a7d)%>%gather(.,a7,value,-study_id)%>%na.omit()%>%filter(value>0)%>%group_by(study_id)%>%summarise(n=n())%>%filter(n>1)%>%pull(study_id)
a7im=a7im%>%filter(!study_id%in%a7imid)
dat_01=left_join(dat_01,a7im,"study_id")%>%dplyr::select(-a7a,-a7b,-a7c,-a7d)

#here put the variables you want!!! #All data
dat_01f=dat_01%>%select(f5,e3,e4,e8,e11:e13,c7,c11,c12,problem_drinking,a5,a6,a8d,a2,g6,g7,g8,g3,g15,g16)%>%na.omit()
nrow(dat_01f)/nrow(dat2001)

#e8 and a7 cause most missing. they both describe the "with whom you current live" can include them if you want.
# All data
gg_miss_upset(dat_01%>%select(f5,e3,e4,c7,c7,c11,c12,problem_drinking,a5,a6,a8d,a2,g6,g7,g8,g3,g15,g16,c4))
gg_miss_upset(dat2001%>%dplyr::select(f5,e3,e4,e11:e13))

```

Variables for Drinking 1997

```{r}

# Use varibales c5,c11, c12, and problem_drinkin for latent variable

dat_97$c11[is.na(dat_97$c11) == TRUE] <- 1
dat_97$c12[is.na(dat_97$c12) == TRUE] <- 1

#Imputing NA's as "None" in c17
dat_97$c16_a[is.na(dat_97$c16_a) == TRUE] <- 1
dat_97$c16_b[is.na(dat_97$c16_b) == TRUE] <- 1
dat_97$c16_c[is.na(dat_97$c16_c) == TRUE] <- 1
dat_97$c16_d[is.na(dat_97$c16_d) == TRUE] <- 1
dat_97$c16_e[is.na(dat_97$c16_e) == TRUE] <- 1
dat_97$c16_f[is.na(dat_97$c16_f) == TRUE] <- 1
dat_97$c16_g[is.na(dat_97$c16_g) == TRUE] <- 1
dat_97$c16_h[is.na(dat_97$c16_h) == TRUE] <- 1
dat_97$c16_i[is.na(dat_97$c16_i) == TRUE] <- 1
dat_97$c16_j[is.na(dat_97$c16_j) == TRUE] <- 1
dat_97$c16_k[is.na(dat_97$c16_k) == TRUE] <- 1
dat_97$c16_l[is.na(dat_97$c16_l) == TRUE] <- 1


#Creating indicators for "3 or more times"
dat_97 <- dat_97 %>%
  mutate(indic_c16a = ifelse(c16_a >=3, 1, 0)) %>% 
  mutate(indic_c16b = ifelse(c16_b >=3, 1, 0)) %>%
  mutate(indic_c16c = ifelse(c16_c >=3, 1, 0)) %>%
  mutate(indic_c16d = ifelse(c16_d >=3, 1, 0)) %>%
  mutate(indic_c16e = ifelse(c16_e >=3, 1, 0)) %>%
  mutate(indic_c16f = ifelse(c16_f >=3, 1, 0)) %>%
  mutate(indic_c16g = ifelse(c16_g >=3, 1, 0)) %>%
  mutate(indic_c16h = ifelse(c16_h >=3, 1, 0)) %>%
  mutate(indic_c16i = ifelse(c16_i >=3, 1, 0)) %>%
  mutate(indic_c16j = ifelse(c16_j >=3, 1, 0)) %>%
  mutate(indic_c16k = ifelse(c16_k >=3, 1, 0)) %>%
  mutate(indic_c16l = ifelse(c16_l >=3, 1, 0)) %>% 
  mutate(problem_drinking = indic_c16a + indic_c16b + indic_c16c + indic_c16d + indic_c16e + indic_c16f + indic_c16g + indic_c16h + indic_c16i +indic_c16j + indic_c16k + indic_c16l)
```



If consider 1997 and 2001:


smooking: (only this one is overlapping)
1997:e3
2001:e4

background(only cosider corr>0.1):
1997:a2,a8,g8,g3,g15,g14,g16
2001:a2,a7,g6,g3,g15,g14,g16
(a8(1997)-a7(2001))

drinking: c5,c11, c12, c16a-l (these correspond directly to 2001, just different number question)(maybe include c8)
greek life: a5, a6, a8d (exact same as 2001)




```{r}

## comparable dataset

# plz add drinking here (BK: I did, just added c7, for 2001)
ids01_97=dat_01%>% dplyr::select(study_id,a5,a6,a8d,c7,c11,c12,problem_drinking,f5,e4,a2,g6,g3,g15,g14,g16)%>%na.omit()%>%pull(study_id)
# select complete cases for 2001 ###overlapping
dat_01j=dat_01%>%filter(study_id%in%ids01_97,f5!=99,f5!=10)%>%select(study_id,a5,a6,a8d,c7,c11,c12,problem_drinking,f5,e4,a2,g6,g3,g15,g14,g16)
nrow(dat_01_97)/nrow(dat2001)

# plz add drinking here (BK: I did, just added c5, for 1997)
ids97_01=dat_97%>% dplyr::select(study_id,e3,f4,a2,a9,c5,c11,c12,problem_drinking,g8,g3,g15,g14,g16,c8)%>%na.omit()%>%pull(study_id)

# select complete cases for 2001 ###overlapping
dat_97j=dat_97%>%filter(study_id%in%ids97_01,f4!=10)%>%select(study_id,e3,f4,a2,a9,c5,c11,c12,problem_drinking,g8,g3,g15,g14,g16)
nrow(dat_97_01)/nrow(dat1997)

```
Define a function to post-process posterior summary
```{r}
summarize_lavaan=function(fit){
  a=summary(fit)
  start=which(a$PE$op=="~") %>% min()
  end=which(a$PE$op=="~") %>% max()
  table=a$PE[start:end,5:8] %>% round(2)
  rownames(table)=paste(a$PE[start:end,1],a$PE[start:end,2],a$PE[start:end,3])
  colnames(table)=c("estimate","SE","z-score","P-value")
  return(table)
}
```

Build a model on 2001-non-overlapping variables (Recommended)
Drink: c7, 11 12
Smoke: e3, 8, 11
Greek: a5,a8d

Background variables used here:
a2: male or not
g6: health
g15: mother drinking tendency
```{r}
model <- '
  # measurement model
     drink=~ c7 + c11 + c12
     smoke=~ e3 + e8 + e11 
     academic=~ f5 
     greek=~ a5 + a8d
     male=~ a2
     health=~ g6 
     mother=~ g15
  # regressions
    academic ~  drink + smoke + greek + male + health
    drink ~ greek + male + mother
    smoke ~ greek + male
  # residual correlations
    c7 ~~ c11 + c12
    e3 ~~ e8 + e11
'
fit_01_non_overlap <- sem(model, data=dat_01f)
table01=summarize_lavaan(fit_01_non_overlap)
```
Look at whether all regression coefficients are significant: academic~greek,  smoke~greek are not significant. Some interpretation can be done here.
```{r}
table01
```


Build a model on 1997-overlapping variables (Recommened)

Academics:
f4
Drinking:
c5, 11 ,12
Smoking:
e3
Greek:
a9

Background variables used here:
a2: male or not
g8: health
g15: mother drinking tendency
```{r}
model <- '
  # measurement model
     drink=~ c5 + c11 + c12
     smoke=~ e3 
     academic=~ f4 
     greek=~ a9
     male=~ a2
     health=~ g8
     mother=~ g15
  # regressions
    academic ~  drink + smoke + greek + male + health
    drink ~ greek + male + mother
    smoke ~ greek + male
  # residual correlations
    c5 ~~ c11 + c12
 
'
fit_97_overlap <- sem(model, data=dat_97j)
table97=summarize_lavaan(fit_97_overlap)
```
```{r}
table97
```


Build a model on 2001-overlapping variables (Not Recommended)

Background variables used here:
a2: male or not
g6: health
g15: mother drinking tendency
```{r}
model <- '
  # measurement model
     drink=~ c7 + c11 + c12
     smoke=~ e4 
     academic=~ f5 
     greek=~ a5 + a8d
     male=~ a2
     health=~ g6 
     mother=~ g15
  # regressions
    academic ~  drink + smoke + greek + male + health
    drink ~ greek + male + mother
    smoke ~ greek + male
  # residual correlations
    c7 ~~ c11 + c12
   
'
fit_01_overlap <- sem(model, data=dat_01j)
summary(fit_01_overlap, standardized=TRUE)
```